# Interactive component
## Data preprocessing for d3 component 

The preprocessing for the interactive component consisted of calculating the number of wildfires per state, per year. We chose the subset of 2013 - 2016 for an interactive component, resulting in the four tables below after preprocessing.

Recall, the NFIRS database includes only a sample of total fire data. As such, rather than interpret the number of wildfires in each state as a raw number, it is more keen to track the number of NFIRS wildfires over time, observing a general trend. 

```{r}
library(lubridate)
library(dplyr)
library(pivottabler)
```

```{r}
# read data
#usdm <- readRDS('./data/usdm.rds')
#wildfires <- readRDS('./data/wildfires.rds')
#wildfires$date = lubridate::mdy(wildfires$INC_DATE) # convert date
#wildfires$year = year(wildfires$date) # pull year from date
```

```{r}
#head(usdm)
```

```{r}
#head(wildfires)
```

```{r}
wildfires2016 <- filter(wildfires, year==2016)
pt <- PivotTable$new()
pt$addData(wildfires2016)

pt$addRowDataGroups("STATE")             #     **** LINE BELOW ADDED ****
pt$defineCalculation(calculationName="count", summariseExpression="n()")
pt$renderPivot()

state_num_wildfire <- pt$asDataFrame()
#write.csv(state_num_wildfire, 'state_num_wildfire2016.csv')

```

```{r}
library(dplyr)
wildfires2015 <- filter(wildfires, year==2015)
library(pivottabler)
pt <- PivotTable$new()
pt$addData(wildfires2015)

pt$addRowDataGroups("STATE")             #     **** LINE BELOW ADDED ****
pt$defineCalculation(calculationName="count", summariseExpression="n()")
pt$renderPivot()

state_num_wildfire <- pt$asDataFrame()
#write.csv(state_num_wildfire, 'state_num_wildfire2015.csv')

```

```{r}
library(dplyr)
wildfires2014 <- filter(wildfires, year==2014)
library(pivottabler)
pt <- PivotTable$new()
pt$addData(wildfires2014)

pt$addRowDataGroups("STATE")             #     **** LINE BELOW ADDED ****
pt$defineCalculation(calculationName="count", summariseExpression="n()")
pt$renderPivot()

state_num_wildfire <- pt$asDataFrame()
#write.csv(state_num_wildfire, 'state_num_wildfire2014.csv')

```

```{r}
library(dplyr)
wildfires2013 <- filter(wildfires, year==2013)
library(pivottabler)
pt <- PivotTable$new()
pt$addData(wildfires2013)

pt$addRowDataGroups("STATE")             #     **** LINE BELOW ADDED ****
pt$defineCalculation(calculationName="count", summariseExpression="n()")
pt$renderPivot()

state_num_wildfire <- pt$asDataFrame()
#write.csv(state_num_wildfire, 'state_num_wildfire2013.csv')

```

## Code for d3 component 
<iframe src="d3-viz.html"
style="position: relative; height: 100%; width: 100%;">
</iframe>


